Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-05-17 trunk/papers/ArticleZ80/source/paper.tex

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{float}
%\usepackage{supertabular} % used in B Symbols
%\geometry{a4paper} % or letterpaper (US) or a5paper or....

@@

The modeling of instruction set, interruptions and input/output ports become the
animation interesting. The animation in ProB is not so fast and practical as a
simulation, for example in \cite{Simulator_z80}, because ProB works manipulating
logic and math concepts that can have infinite sets and non determinism, then
ProB requires more processing, specially when it works analysing  complex
expressions or very large expressions. However, a user of ProB with this
modeling can: execute manually a sequence of Z80's instructions that represent a
program and view the visited states of ports, memory and registers. It is
important to debug the effect action on the state space computed; create a
tracing of the states; analyse constants, functions and expressions; and search
possible violations.

\subsection{Verification process}% ESTOU TRABALHANDO AQUI
% OK -> como as regras foram utilizadas ?  Respondido na seção das regras
% TODO:ENGLISH REVIEW parágrafo inicial: VISÃO GERAL
This project has many proof obligations and its verification can be a hard task,
hence this section explains about the verification process of proof
obligations and its importance. % [O que as provas garantem? citar que énecessário buscar técnicas eficientes para realizar as provas ]
The proof obligations allow to verify the data types, important system properties and if
the expressions are well-defined (WD)\footnote{An expression is called ``well-defined''
(or unambiguous) if its definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees because they can set many safety
rules. However, the model can be very difficult to prove.

 % -> O que significa cada coluna e linha da tabela ?
The Table~\ref{tab:completestatistcs} contains some statistics of verification
process, it shows only the number of no obvious proof
obligations\footnote{Obvious proof obligations are simple formulas automatically eliminated by
the tool.} and the titles of columns use acronyms: {\small \textbf{PO}} - number
of proof obligations; {\small \textbf{UPO}} - number of unproved proof
obligations; {\small \textbf{WD PO}} - number of proof obligations  related to
well definition lemmas; {\small \textbf{UWD PO}} - number of unproved proof
obligations related to well definition lemmas.

@@

% Porque o modelo não foi verificado completamente ?
There are many benefits when a formal model is developed, but to verify the
model is not an easy task, the first version of Z80 model have already verified
completely. However, this new Z80 model with support to animation in ProB has not
been verified completely because the changes in Z80 model are recent and the
verification time is long.
% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
Nevertheless this project has 3631 proof obligations and the first version of
Z80 model spent approximately two months of work to finish the verification
process. The same  developed verification strategies were adjusted and reused in
new Z80 model, so approximately  7 days of work was enough to verify 3062  proof
obligations (84\%) and remaining only 569 proof obligations (16\%) not yet
verified.

Two development strategies were important to improve the verification of 3062
proof obligations: the intensive use of parallelism and user pass. The Z80 module
was decomposed in 8 modules of instructions, it was very helpful to decrease the
time for verification. Because AtelierB can run simultaneously \textit{N} instances of 
provers, one per module, reducing the time verification by up to \textit{N}  times when
a processor with \textit{N} cores is used  or the process is submitted to \textit{N} 
networked computers. So we also set up a proving environment consisting of networked
computers to take advantage of the distribution facilities provided in the B development
environment.
% Resumo de como aconteceu o processo de verificação,
A second interesting strategy  was to reuse the user pass. Usually, an used
user pass to aid the verification of one proof obligation can solve several
similar proof obligations. Thus, few generic user pass are needed to prove many
proof obligations. As there are many similar assembly instructions, some
human-directed proofs, when replayed, could discharge other proof obligations. A
good example is a set of 17 user pass that quickly aided the verification of
approximately 2533 of WD proofs.

@@

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. 
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful %TODO: COLOCAR APÓS TODOS OS BENEFICIOS PROPORCIONADOS COM A SIMULAÇÃO e COMPATIBILIDADE COM O PROB. (OBTER DO PARAGRAFO ANTERIOR DA SEÇAO VERIFICATION PROCESS).
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level.

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object from the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

%\section{Case study}
%\include{Petroleumproductiontestsystem}

>>>>

%%% PAGE DIMENSIONS
%\usepackage{geometry} % to change the page dimensions
\usepackage{graphicx,array}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{epsfig}
\usepackage{float}
%\usepackage{supertabular} % used in B Symbols
%\geometry{a4paper} % or letterpaper (US) or a5paper or....

@@

\subsection{Verification and validation}% ESTOU TRABALHANDO AQUI
% OK -> como as regras foram utilizadas ?  Respondido na seção das regras
% TODO:ENGLISH REVIEW parágrafo inicial: VISÃO GERAL

The model presented in this paper is suitable both for validation
through animation with ProB, and for verification through the standard
proof obligations generated by the application of the B method.

The animation in ProB is not as fast and practical as a simulation
with a dedicated tool (e.g. \cite{Simulator_z80}). ProB is a
general-purpose animator of formal specification, and addresses
computationally more difficult problems. Anyway, it is capable of
animating, and also possibly model checking, a formal model of the
instruction set of the Z80, which was our initial motivation in
constructing that model. One may simulate step-by-step sequences of
Z80 instructions and visualize the visited states of ports, memory and
registers. It is a useful tool to debug by creating a trace of the
visited states, offering to analyse constants, functions and
expressions, and search possible violations.

The construction of the model with B generates a large number of proof
obligations and the verification is a time-consuming task. We describe
here some details on this effort and some solutions we have devised to
reduce the effort.  In B, the generated proof obligations allow to
verify the data satisfy invariant properties, check important system
properties and that the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}.

Table~\ref{tab:completestatistcs} displays statistics of the
verification process. Note that the figures in it include only the
number of \emph{non obvious} proof obligations\footnote{Obvious proof
  obligations are simple formulas automatically eliminated by the
  tool.}. The titles of columns use acronyms: {\small \textbf{PO}} -
number of proof obligations; {\small \textbf{UPO}} - number of (yet)
unproved proof obligations; {\small \textbf{WD PO}} - number of proof
obligations related to well definition lemmas; {\small \textbf{UWD
    PO}} - number of (yet) unproved proof obligations related to well
definition lemmas.

@@

% Porque o modelo não foi verificado completamente ?
There is still a large number of open proof obligations, i.e. we have
not yet found their proof, or they are not valid and our model is not
correct. Each such proof obligation requires interaction with a proof
assistant. We see that for a project of that size, a large effort
needs to be dedicated to find those proofs. One interesting feature of
the proof assistant is, once a proof has been found, it can
automatically apply it to other proof obligations, as it is often the
case that the same proof script discharges several proof obligations.

% ->Quanto tempo foi gasto nessa verificação ?  Aproximadamente quanto tempo de
% processamento elas reduziram ?
We had a previous experience with a similar effot, where we were able
to discharge all the proof obligations: that of the model of the Z80
instruction set designed without restricting ourselves to constructs
that may be animated with ProB, and without modeling external actions
(as described in section~\ref{sec:externalactions}). In that first
experience, the verification took us about two months. In that second
project, we have been able to verify 3062 (e.g. roughly 80 \%) of 3631
proof obligations in one week.

In order to be able to verify such a large number of proof obligations
in such a short term, we employed two strategies: parallelism and
\emph{user pass}. AtelierB has an option to run any number of threads
in parallel and also to send proof obligations to other, possibly
remote, processes running AtelierB. With those options we had the
possibility to have eight concurrent instances of the prover. We thus
split the instruction set of the Z80 in eight sub-modules and checked
them concurrently (each instance of AtelierB may only verify one module
at a time).

The second strategy, called \emph{user pass}, takes advantage of
another feature of the AtelierB prover. We defined a number of proof
scripts which are then applied to all remaining proof obligations. Few
generic user passes are enough to prove many proof obligations. Also,
as the models of many instructionsa are similar, proofs found in
interactive mode may similarly be replayed and discharge other proof
obligations. We have defined a set of 17 user passes that are able
to discharge more than 2500 WD proof obliagations.

@@

The Z80 formal model provides many benefits, because of the
verification of generated proof obligations guarantees: the correct
use of data types, the developed security properties and the
well-defined of all the expressions.  Furthermore, the designer has a
big flexibility to create new and specific security properties, this
is very
useful %TODO: COLOCAR APÓS TODOS OS BENEFICIOS PROPORCIONADOS COM A SIMULAÇÃO e COMPATIBILIDADE COM O PROB. (OBTER DO PARAGRAFO ANTERIOR DA SEÇAO VERIFICATION PROCESS).
to adjust the verification in accordance with the
requirements. Moreover, this example of model could complement the
existing documentation for users and assembly programmers.

But our goal is to use the model of the Z80 instruction set to extend
the scope of the B method up to the assembly
level~\cite{Dantas_SBMF08}. Indeed, we realized a case study where an
abstract B model had been refined up to the algorithmic model using
the classic development strategy. The algorithmic model was manually
compiled to a Z80 assembly program. We then constructed a B module
composed with the instructions of that program, and importing the
formal model of the Z80 instruction set. We then showed that this
assembly-level module was indeed a refinement of the original abstract
model.

This case study is a component of a pilot project to analyse a
petroleum production test system. Its modeling was developed according
to \cite{LAUT_SERGIO} and its B assembly model could also be animated
with ProB.
